Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-24 paper/paper.tex
Add more details to Section 6 (beta eta)

\section{Existence of $\beta\eta$-normal forms}
\scw{relocated text from previous section. Not a complete description yet.}

We first extend parallel reduction to include $\eta$ reduction for functions.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}

\Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
equational theory since convertibility ($[[a <=> b]]$), the untyped relation
used for type conversion, is built on top of parallel reduction. We can
recover the same confluence result about parallel reduction using the standard
techniques from \citet{barendregt:lambda-calculi-with-types,
  takahashi-parallel-reduction}, though specifically for the de Bruijn
representation, we need the following anti-renaming lemma about parallel
reduction.

Both the confluence and anti-renaming properties require structural
induction over the size of terms to obtain a useful induction
hypotehsis in the $\eta$ case.

This lemma is mainly to show that \rref{P-AbsEta} does not introduce
any new $\beta$ redex because $\eta$ reduction is the only real step
we can take.

\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}
\begin{proof}
  From Corollary~\ref{corollary:exbetanf}, we can easily conclude the
  existence of $\beta\eta$-normal forms for well-typed terms since
  $\eta$-reduction preserves $\beta$-normal form (implied by
  Lemma~\ref{lemma:parnenf}) and also strictly decreases the size of
  the term (evident from \rref{P-AbsEta}).
\end{proof}

@@

\subsection{Extraction of conversion algorithm}
\yl{technically future work. Not sure whereelse to put it}

>>>>

\section{Existence of $\beta\eta$-normal forms}
We can easily incorporate the function $\eta$ law to the equational
theory of \lang{} by adding the following parallel reduction rule.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
% \Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
% equational theory since convertibility ($[[a <=> b]]$), the untyped relation
% used for type conversion, is built on top of parallel reduction.
We can recover the same confluence result about parallel reduction
using the standard techniques from
\citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though anti-renaming
(Lemma~\ref{lemma:parantirenaming}) needs to be proven before the
diamond property (Lemma~\ref{lemma:pardiamond}). Another complication
is that confluence and anti-renaming now requires induction on a size
metric of lambda terms so the $\eta$ case has a useful induction hypothesis.

Note that the specification of our logical relation does not require
any changes. We can continue using $\ottkw{ne}$ and $\ottkw{nf}$ to
represent $\beta$-neutral and normal forms. By reproving the
fundamental lemma, we learn that every well-typed term as a
$\beta$-normal form. Lemma~\ref{lemma:parnenf} still holds with our
extended parallel reduction. In the presence of the $\eta$ reduction
rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
preserves $\beta$-normal forms (i.e. does not produce new
$\beta$-redexes). Furthermore, since the $\eta$ reduction rule for
functions strictly decreases the size of the term, the existence of
$\beta\eta$ normal form trivially follows.
\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}
Due to some technical
details, our proof about the existence of $\beta\eta$ form does not
immediately induce a decision procedure for type conversion in
\lang{}. We discuss this issue and its workaround in
Section~\ref{sec:conversionalgo}.

@@

\subsection{Extraction of conversion algorithm}
\label{sec:conversionalgo}
\yl{technically future work. Not sure whereelse to put it}
